Network based multifactorial modelling microRNA-target interaction.

### 2019-05-27 article.Rmd
abstract edited, into first paragraph edited

Competing endogeneous RNA (ceRNA) regulations and crosstalk in human are remarkable issue in mean of regulations via miRNAs. Many studies have pointed out that an alteration in miRNA:target interaction can result in unexpected regulations because of indirect and complex interaction. In this paper, we defined a new network-based model that handles miRNA:ceRNA interactions with expression values and interaction factors optionally. We applied approach in sample dataset and defined dynamics of it. We have observed that a perturbation in an element of network affects whole competing elements differently and cooperative efficiencies of miRNAs on common targets could be shown. Importance of miRNA:target ratios was also observed consistently with previous reports. We have obtained that the competing elements which have the same or close expression values may not affected equal from the perturbation because of repression functionality of interactions. To apply model to the real sample that obtained combining  breeast cancer gene and miRNA expression dataset via high throughput experimental miRNA:target dataset, we selected a gene, *ABCC1*, which over expressed in tumor tissue and utilized it as perturbating element. We have observed that one change can be sufficient to perturb regulations, and may not predicted all of regulations correctly, with this change. Therefore, this model which helps to understand crosstalking between elements could be tested and improved via results of well-constructed experimental dataset and takes to hope for future research.

@@

MicroRNAs (miRNAs) that are key regulator of gene expression through various post-transcriptional mechanisms are a family of short non-coding RNAs. Although repression mechanisms of miRNA are not fully understood, exact effect of miRNAs is regulation of direct target/targets negatively. Repressive activities of miRNAs vary depending on many factors that are significant to microRNA:target interactions. These factors include microRNA:target binding energy, binding location in target sequence, base pairing types between miRNA and target, abundance of miRNAs and targets [@grimson_microrna_2007]. Binding energies of miRNA:target complexes vary based on nucleotide context and determine folding stability of complex [@cao_predicting_2012]. It has been demonstrated that the binding energy between miRNA and target indicates stability or affinity of complex [@helwak_mapping_2013] and does not directly determine repressive activity of miRNA [@cao_predicting_2012]. Early studies have argued that 2-8 nt sequence, seed, located in miRNA 5’UTR bind to specific sequence located in 3’UTR of its target [@bartel_micrornas_2004; @lewis_conserved_2005]. In recent studies, it has been shown miRNAs can interact with targets via sequence that located in different region such as 5’UTR or CDS [@hausser_analysis_2013; @helwak_mapping_2013; @moore_mirnatarget_2015]. These studies also shown that binding location could indicate functionality of miRNA:target interaction or affect profile of target abundance. It has been shown that miRNAs exhibit repressive activity via, 6-8 nt long sequence that is prefectly complementary with targets, seed region at the 5’UTR of miRNAs [@bartel_micrornas:_2009; @grimson_microrna_2007]. On the other hand, some researchers have reported that seed sequence of miRNA can have mismatches or bulged/wobble nucleotides and locates in different region of miRNAs [@hafner_transcriptome-wide_2010; @helwak_mapping_2013]. In addition to all of these, abundance of miRNAs and targets and miRNA:target ratio in cells are determinant for efficiency of miRNA:target interaction [@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014].

As it is possible for miRNAs to suppress multiple targets, an individual RNA molecule can also be targeted by various miRNAs. In that case, the targeted RNAs exhibit competitor behavior, that is hypothesized as competing endogenous RNAs (ceRNAs) [@ala_integrated_2013; @cesana_deciphering_2013], against their miRNAs. Briefly, Ala et al. have explained the ceRNA hypothesis as disturbtion of the other target when a steady-state system that included one miRNA and two target, was perturbed with expression changes at one of the targets [@ala_integrated_2013]. When interaction between miRNAs and their target were dealed in the cell, explanation and prediction of results of perturbation is difficult due to varieties of interactions. Various computational and experimental studies have applied to be understood for ceRNA:miRNA interactions. It has been observed that when abundance of one of the targets of miR-122 was increased, the other target expression also slightly increased as a result of decreasing repressive activity of miR-122 on them [@denzler_assessing_2014]. Bosson et al. have developed a mathematical model for changes on total target pool concentration after grouping to affinity of targets. They have demonstrated that miRNA activity correlated with affinity between miRNA and target [@bosson_endogenous_2014]. Cooperative efficiency of miRNAs as well as competitor behaviors of targets also were studied and it has been demonstrated crucial for regulations at expression of targets [@denzler_impact_2016]. MiRNA:target interactions have modelled as stoichiometric and catalytic mechanisms and Figliuzzi et al. have recommended to handling models in network structures [@figliuzzi_micrornas_2013]. The model that can explain miRNA target interaction through topological features has been applied at bipartite network by Nitzan et al. [@nitzan_interactions_2014]. Robinson and Henderson applied the model that handles miRNA:target direct and indirect interactions via common miRNA of gene and target of miRNA, at bipartite network. It has been demonstrated that whole miRNAs and targets in network can interacted through common miRNAs and genes [@robinson_modelling_2018]. Associated genes that are targets of the same miRNAs have been found with help of correlation of gene expression changes in recent study [@markus_list_sponge_2017]. List et al. have specified that their approach can be useful for ceRNA’s studies and published these approach as a package that can use with R programming.

>>>>

Competing endogenous RNA (ceRNA) regulations and crosstalk between various types of non-coding RNA in human is remarkable in means of miRNA regulation. Many studies have pointed out that an alteration in miRNA:target interaction can result in unexpected changes due to indirect and complex interactions. In this paper, we defined a new network-based model that handles miRNA:ceRNA interactions with expression values. Our model is able to handle miRNA interaction factors such as seed type, binding energy, if provided. Our approach is able to reveal that a perturbation in an element of network affects whole competing elements differently and cooperative efficiencies of miRNAs on common targets could be calculated. Our findings emphasized importance of miRNA:target ratios being crucial, as reported by previous studies. We have showed that the competing elements which have the same or close expression values may not be affected equally from the perturbation because of repression functionality of interactions(TOFIX son kısmı anlayamadım). We applied the model to real sample consisting of breast cancer gene and miRNA expression dataset and experimental miRNA:target interaction dataset all generated via high throughput sequencing methods. A gene over-expressed in tumor tissue, namely *ABCC1*, is used as perturbing element. We have observed that change in expression level of single gene in miRNA:target network is sufficient to perturb regulations in whole network, due to unforeseen and unpredicted regulation which are only visible when considered in network context. Therefore, this model helps unveiling the crosstalk between elements in miRNA:target network where abundance of target and sponge effect are taken into account. The model is scalable and can be plugged in with emerging miRNA effectors such as circRNAs. The model is available as R package at https://github.com/selcenari/ceRNAnetsim. (TODO the link will change)

@@

MicroRNAs (miRNAs) are a family of short non-coding RNAs which are key regulator of gene expression through various post-transcriptional mechanisms. Although the mechanisms by which miRNA represses are not fully understood,  miRNAs predominantly repress their targets. Repressive activities of miRNAs vary depending on many factors that are significant to microRNA:target interactions. These factors include microRNA:target binding energy, binding location in target sequence, base pairing types between miRNA and target, abundance of miRNAs and targets [@grimson_microrna_2007]. Binding energies of miRNA:target complexes vary based on nucleotide context and determine folding stability of complex [@cao_predicting_2012]. It has been demonstrated that the binding energy between miRNA and target indicates stability or affinity of complex [@helwak_mapping_2013] and does not directly determine repressive activity of miRNA [@cao_predicting_2012]. Early studies have argued that 2-8 nt sequence, seed, located in miRNA 5’UTR (TOFIX mirna için UTR ifadesi geçerli mi, sadec 5 ucu dense yeterli mi?) bind to specific sequence located in 3’UTR of its target [@bartel_micrornas_2004; @lewis_conserved_2005]. In recent studies, it has been shown that miRNAs can interact with targets via sequences located in regions such as 5’UTR or CDS [@hausser_analysis_2013; @helwak_mapping_2013; @moore_mirnatarget_2015]. These studies also showed that binding location could indicate functionality of miRNA:target interaction or affect profile of target abundance(TOFIX abundance etkisi derken, bazen azaltabilir, bazen arttırabilir, manasında mı?). It has been shown that miRNAs exhibit repressive activity via, 6-8 nt long sequence that is perfectly complementary with targets, seed region at the 5’UTR of miRNAs(TOFIX mirna için UTR geçerli mi) [@bartel_micrornas:_2009; @grimson_microrna_2007]. On the other hand, some researchers have reported that seed sequence of miRNA can have mismatches or bulged/wobble nucleotides and may locate in region other than 5'end of miRNAs [@hafner_transcriptome-wide_2010; @helwak_mapping_2013]. On top of all these factors, abundance of miRNAs and targets and miRNA:target ratio in cells predominantly affect efficiency of miRNA:target interaction [@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014].

As it is possible for miRNAs to suppress multiple targets, an individual mRNA molecule can also be targeted by multiple miRNAs. In that case, the targeted mRNAs exhibit competitor behavior, that is hypothesized as competing endogenous RNAs (ceRNAs) [@ala_integrated_2013; @cesana_deciphering_2013], against their miRNAs. Briefly, Ala et al. have explained the ceRNA hypothesis as distribution of the other target when a steady-state system that included one miRNA and two target, was perturbed with expression changes at one of the targets [@ala_integrated_2013]. When interaction between miRNAs and their target were dealed in the cell, explanation and prediction of results of perturbation is difficult due to varieties of interactions. Various computational and experimental studies have applied to be understood for ceRNA:miRNA interactions. It has been observed that when abundance of one of the targets of miR-122 was increased, the other target expression also slightly increased as a result of decreasing repressive activity of miR-122 on them [@denzler_assessing_2014]. Bosson et al. have developed a mathematical model for changes on total target pool concentration after grouping to affinity of targets. They have demonstrated that miRNA activity correlated with affinity between miRNA and target [@bosson_endogenous_2014]. Cooperative efficiency of miRNAs as well as competitor behaviors of targets also were studied and it has been demonstrated crucial for regulations at expression of targets [@denzler_impact_2016]. MiRNA:target interactions have modeled as stoichiometric and catalytic mechanisms and Figliuzzi et al. have recommended to handling models in network structures [@figliuzzi_micrornas_2013]. The model that can explain miRNA target interaction through topological features has been applied at bipartite network by Nitzan et al. [@nitzan_interactions_2014]. Robinson and Henderson applied the model that handles miRNA:target direct and indirect interactions via common miRNA of gene and target of miRNA, at bipartite network. It has been demonstrated that whole miRNAs and targets in network can interacted through common miRNAs and genes [@robinson_modelling_2018]. Associated genes that are targets of the same miRNAs have been found with help of correlation of gene expression changes in recent study [@markus_list_sponge_2017]. List et al. have specified that their approach can be useful for ceRNA’s studies and published these approach as a package that can use with R programming.
